---
title: "Crafted Pizza - Data Cleaning"
author: "John Mwangi"
date: "1/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading required packages

```{r}
rm(list = ls())

if(!require(pacman)){install.packages("pacman")}
pacman::p_load(readxl, writexl, scales, lubridate, tidyverse)
```

Setting folder paths

```{r}
# Working dir
getwd()

# Folder paths
sales_path <- "../../../RawData/Sales/"
expenses_path <- "../../../RawData/Expenses/"
main_path <- "../../../Main/"

# Branches
sheets <- c("KK", "KB")

# Pizza price lists
pizza_large <- c(850,900,950,1000)
pizza_medium <- c(650,700,750,800)
pizza_small <- c(450,500,550,600)
```

Read all sheets in one excel and combine them.

```{r}
tryCatch(expr = read_excel(path = "../../../RawData/Sales/Sep2021.xlsx", sheet = "KK"),
         error = function(e){
           read_excel(path = "../../../RawData/Sales/Sep2021.xlsx")
         })

try(expr = read_excel(path = "../../../RawData/Sep2021.xlsx", sheet = "KB"),
    silent = TRUE)
```


```{r}
read_sales <- function(filepaths){
  
  file_kk <- tryCatch(expr = read_excel(path = filepaths, 
                                        sheet = sheets[1],
                                        col_names = FALSE) %>% 
                        set_names("data") %>% 
                        mutate(filepath = filepaths),
                      
         error = function(e){
           read_excel(path = filepaths,
                      col_names = FALSE) %>%
             set_names("data") %>% 
             mutate(filepath = filepaths)
         })
  
  file_kk <- file_kk %>% mutate(branch = sheets[1])

  file_kb <- try(expr = read_excel(path = filepaths, 
                                   sheet = sheets[2], 
                                   col_names = FALSE) %>% 
                   set_names("data") %>% 
                   mutate(filepath = filepaths,
                          branch = sheets[2]),
                 silent = TRUE)
  
  file_df <- rbind(file_kk, file_kb)
  
  file_df <- file_df %>% filter(str_detect(string = filepath, pattern = "xlsx$"))
  
  return(file_df)
}

read_sales(filepaths = "../../../RawData/Sales/Sep2021.xlsx")
```

Read all files in one go.

```{r}
filepaths <- 
  dir(path = sales_path, 
      all.files = FALSE, 
      pattern = ".xlsx$", 
      full.names = TRUE)

filepaths

sales_data <- filepaths %>% 
  map_dfr(read_sales) %>% 
  mutate(data = str_to_lower(string = data))

sales_data
```

Extract branch

```{r}
sales_data <-
sales_data %>% 
  mutate(branch = if_else(branch=="KK", true = "kikuyu", false = "kabete"))

sales_data %>% count(branch)
```

Extract year and month. Create date column.


```{r}
sales_data <- 
  sales_data %>% 
  extract(col = filepath, 
          into = c("month","year"), 
          regex = ".*/([[:alpha:]]+)([[:digit:]]+)")


sales_data <- 
  sales_data %>% 
  mutate(day = str_extract(string = data, pattern = "^\\d\\d?\\w\\w")) %>% 
  fill(day, .direction = "down") %>% 
  mutate(day = parse_number(day),
         date = paste0(day,month,year),
         date = dmy(date)) %>% 
  select(sales = data, branch, date)

sales_data <-
sales_data %>% 
  arrange(desc(date),
          branch)

sales_data %>% group_by(branch) %>% summarise(range(date))
```

Extract product, price and location.

```{r}
sales_data <-
sales_data %>% 
  extract(col = sales, into = c("product","price_"), regex = "([[:alpha:]]+.*)-(\\d.*)", remove = FALSE, convert = FALSE) %>% 
  mutate(price_ = str_remove_all(string = price_, pattern = "/[[:alpha:]]+")) %>% 
  separate(col = price_, into = c("price", "extras"), sep = "/", remove = FALSE, convert = FALSE) %>% 
  mutate(across(.cols = c(price, extras), .fns = parse_number)) %>% 
  select(-price_) %>% 
  extract(col = sales, into = "location", regex = "\\d+-([[:alpha:]]+/?[[:alpha:]]+)", remove = FALSE, convert = FALSE) %>% 
  mutate(location = str_remove(string = location, pattern = "/[[:alpha:]]+"))

sales_data
```

Filter out unnecessary rows i.e. rows without price information but keeping rows with no sales. Replace NAs with zero.

```{r}
sales_data <- sales_data %>% 
  filter(!is.na(price) | !str_detect(sales, "[:digit:]")) %>% 
  replace_na(list(price = 0, extras = 0))

sales_data
```

Include product categories: pizza, sausage, fries, chicken, sandwich,  burgers

```{r}
sausage_regex <- "sausage"
chicken_regex <- "^chicken$"
sandwich_regex <- "sandwich"
fries_regex <- "fries|fry|poussin|masala|cheesy|poisson|^spicy$|spicy f.*"
burger_regex <- "burger|/|bacon"
bhajia_regex <- "bhajia"

sales_data <- sales_data %>% 
  mutate(category = case_when(str_detect(product,sausage_regex) ~ "sausage",
                              str_detect(product,chicken_regex) ~ "chicken",
                              str_detect(product,sandwich_regex) ~ "sandwich",
                              str_detect(product,fries_regex) ~ "fries",
                              str_detect(product,burger_regex) ~ "burger",
                              str_detect(product,bhajia_regex) ~ "bhajia",
                              str_detect(product,"sales") ~ "no sales",
                              TRUE ~ "pizza")) %>% 
  mutate(category = na_if(category,"NA"))


sales_data
```

Corrected spellings. We have varieties of pizza, burgers, sandwiches, fries but not sausages, chicken

```{r}
bce_regex <- "b/e/c|b/c/e|e/c/b|c/e/b|b, c and e|bacon.*egg"
bc_regex <- "b/c burger|bacon and cheese|cheese and bacon|b and c|c and b|bacon cheese|bacon burger and cheese"
md_regex <- "^deluxe$|meat deluxe|meaty deluxe"
poussin_regex <- "poussin|poisson"
regular_fries <- "^fry$|^fries$|salted"
ss_regex <- "sweet|s and"
margh_regex <- "margharita|margarita"
original_burger <- "^burgers?$|original"


sales_data <- 
  sales_data %>% 
  mutate(product_name = case_when(category=="sausage" ~ "sausage",
                                   category=="chicken" ~ "chicken",
                                   TRUE ~ product)) %>% 
  mutate(product_name = case_when(category=="fries" & str_detect(product,"cheesy") ~ "cheesy fries",
                                  category=="fries" & str_detect(product,"masala") ~ "masala fries",
                                  category=="fries" & str_detect(product,poussin_regex) ~ "poussin fries",
                                  category=="fries" & str_detect(product,"spicy") ~  "spicy fries",
                                  category=="fries" & str_detect(product,regular_fries) ~ "regular fries",
                                  TRUE ~ product_name)) %>% 
  mutate(product_name = case_when(category=="sandwich" & str_detect(product,"chicken") ~ "chicken sandwich",
                                   category=="sandwich" & str_detect(product,"beef") ~ "beef sandwich",
                                   category=="sandwich" & str_detect(product,"egg") ~ "egg sandwich",
                                   TRUE ~ product_name)) %>% 
  mutate(product_name = case_when(category=="pizza" & str_detect(product,"periperi") ~ "chicken periperi",
                                   category=="pizza" & str_detect(product,"bbq") ~ "bbq beef",
                                   category=="pizza" & str_detect(product,ss_regex) ~ "sweet and sour chicken",
                                   category=="pizza" & str_detect(product,margh_regex) ~ "margherita",
                                   category=="pizza" & str_detect(product,"hawaian") ~ "hawaiian",
                                   category=="pizza" & str_detect(product,md_regex) ~ "meat deluxe",
                                   category=="pizza" & str_detect(product,"^pepperoni$") ~ "beef pepperoni",
                                   TRUE ~ product_name)) %>%
  mutate(product_name = case_when(category=="burger" & str_detect(product,original_burger) ~ "original burger",
                                   category=="burger" & str_detect(product,bce_regex) ~ "bacon,cheese,egg burger",
                                   category=="burger" & str_detect(product,bc_regex) ~ "bacon and cheese burger",
                                   category=="burger" & str_detect(product,"bacon burger") ~ "bacon burger",
                                   category=="burger" & str_detect(product,"cheese burger") ~ "cheese burger",
                                   TRUE ~ product_name)) %>% 
  mutate(category = ifelse(product_name=="hawaiian/pepperoni","pizza",category))

sales_data
```

Determining product sizes for chicken and pizza

```{r}
# Modulo function. We'll use apply instead of a second for loop over the price vector.
# The idea is to give it a price and return a vector of results then take that and do apply()
mod_price <- function(x, price_list){
  mod <- c()
  for(p in seq_along(price_list)){
    mod[p] <- x %% price_list[p]
  }
  return(mod)
}

# Applying the custom function
sales_data <- 
  sales_data %>% 
  mutate(mod_pizza_large = map(.x = price, .f = mod_price, price_list = pizza_large),
         mod_pizza_medium = map(.x = price, .f = mod_price, price_list = pizza_medium),
         mod_pizza_small = map(.x = price, .f = mod_price, price_list = pizza_small)) %>% 
  mutate(mod_pizza_large = map_dbl(.x = mod_pizza_large, .f = min),
         mod_pizza_medium = map_dbl(.x = mod_pizza_medium, .f = min),
         mod_pizza_small = map_dbl(.x = mod_pizza_small, .f = min)) %>% 
  mutate(size = case_when(mod_pizza_large==0 ~ "large",
                          mod_pizza_medium==0 ~ "medium",
                          mod_pizza_small==0 ~ "small",
                          TRUE ~ "Unknown")) %>% 
  mutate(size = case_when(size=="Unknown" & category=="pizza" & price>=min(pizza_large) ~ "large",
                          size=="Unknown" & category=="pizza" & price<=min(pizza_medium) ~ "small",
                          size=="Unknown" & category=="pizza" & price>min(pizza_medium) & price<min(pizza_large) ~ "medium",
                          TRUE ~ size)) %>% 
  select(-starts_with("mod")) %>% 
  mutate(size = if_else(category == "pizza", true = size, false = "Unknown"))

sales_data

# Alternative approach
# Ref: https://www.youtube.com/watch?v=5lJX4IgmqRA
# mod_pizza_large <- lapply(X = price, FUN = mod_price, price_list = pizza_large)
# mod_pizza <- sapply(X = mod_pizza_large, FUN = function(x) min(unlist(x)))
```

Adding chicken sizes

```{r}
chicken_price <- c(200)

sales_data <- sales_data %>% 
  mutate(size = ifelse(test = category=="chicken", 
                       yes = paste0((price/chicken_price)*0.25,"kg"), 
                       no = size))

sales_data %>% filter(category=="chicken")
```

Everything else the size is same as category.

```{r}
sales_data <- 
  sales_data %>% 
  mutate(size = ifelse(!category %in% c("pizza","chicken"), category, size))

sales_data
```

Final renaming, rearranging, selection of columns

```{r}
colnames(sales_data) <- str_to_title(colnames(sales_data))

sales_data <- sales_data %>%
  select(-Sales,-Product) %>% 
  rename(Pizza = Product_name) %>% 
  mutate(Revenue = Price + Extras) %>% 
  select(Date,Branch,Pizza,Size,Price,Extras,Revenue,Location,Category) %>% 
  mutate_if(.predicate = is.character, .funs = str_to_title)

sales_data
```

Simple summaries for sales.

```{r}
sales_summary <-
sales_data %>% 
  group_by(Branch, Year = year(Date)) %>% 
  mutate(revenue_yr = sum(Revenue, na.rm = TRUE)) %>% 
  ungroup() %>% 
  group_by(Branch, Year, Month = month(Date)) %>% 
  mutate(revenue_mon = sum(Revenue, na.rm = TRUE)) %>% 
  ungroup() %>% 
  select(Branch, Year, Month, revenue_mon, revenue_yr) %>% 
  group_by(Branch, Year, Month) %>% 
  summarise(revenue_mon = mean(revenue_mon),
            revenue_yr = mean(revenue_yr)) %>% 
  arrange(Branch, Year, Month) %>% 
  ungroup()

sales_summary %>% 
  arrange(Branch, desc(Year), Month)
```


```{r}
sales_summary_chart <- sales_summary %>%
  group_by(Branch) %>% 
  mutate(mon_abr = month(x = paste0(Year,"-",Month,"-01"), label = TRUE, abbr = TRUE)) %>% 
  mutate(year_mon = paste0(mon_abr," ",Year)) %>% 
  mutate(period = row_number()) %>% 
  mutate(year_mon = fct_reorder(year_mon, period)) %>% 
  ungroup()

sales_summary_chart
```

```{r}
branches = unique(sales_summary_chart$Branch)

calc_trends <- function(tbl, yr){
  
  trends = tibble(.rows = 0)
  
  for (branch in seq_along(branches)){
    trend_branch <- lm(formula = revenue_mon ~ period, data = tbl %>% filter(Branch==branches[branch] & Year %in% yr)) %>% 
    broom::tidy() %>% 
    mutate(sign = sign(estimate),
           sign = ifelse(test = sign==1,yes = "increasing",no = "decreasing"),
           estimate = round(estimate)) %>% 
    filter(term == "period") %>%
    mutate(branch = branches[branch]) %>% 
    select(branch, estimate, sign)
    
    trends <- bind_rows(trends, trend_branch)
  }
  
  trends <- trends %>% arrange(branch)
  return(trends)
}


calc_trends(tbl = sales_summary_chart, yr = c(2021))
```


```{r}
sales_summary_chart %>% 
  filter(Branch == "Kikuyu" & Year==2021) %>% 
  mutate(year_mon = fct_reorder(year_mon, period)) %>% 
  ggplot(aes(x = year_mon, y = revenue_mon, group = 1)) +
  geom_line() +
  scale_y_continuous(labels = dollar_format(prefix = "KES "))+
  geom_smooth(method = "lm", formula = "y~x")+
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Kikuyu Branch Sales",
       y = NULL,
       x = NULL)

sales_summary_chart %>% 
  group_by(Branch) %>% 
  filter(Year==2021) %>% 
  summarise(sd = sd(revenue_mon))
```

Sales Chart.

```{r}
rev_totals <-
  sales_summary_chart %>% 
  group_by(Branch, Year) %>% 
  summarise(revenue = sum(revenue_mon)) %>% 
  arrange(Branch, Year)

rev_totals
```


```{r}
yr <- c(2021)
trends <- calc_trends(tbl = sales_summary_chart, yr = yr)
rev <- rev_totals %>% filter(Year == yr)

sales_summary_chart %>% 
  filter(Year %in% yr) %>% 
  mutate(year_mon = tidytext::reorder_within(x = year_mon, by = period, within = Branch)) %>% 
  ggplot(aes(x = year_mon, y = revenue_mon, group = 1)) +
  geom_line() +
  scale_y_continuous(labels = dollar_format(prefix = "KES ")) +
  geom_smooth(method = "lm", formula = "y~x") +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~Branch, scales = "free_x") +
  tidytext::scale_x_reordered() +
  labs(title = glue::glue("Monthly Branch Sales in {yr}"),
       subtitle = glue::glue("{trends$branch[1]}: {rev$revenue[1]} {trends$sign[1]} at {trends$estimate[1]}
                             {trends$branch[2]}: {rev$revenue[2]} {trends$sign[2]} at {trends$estimate[2]}"),
       y = NULL,
       x = NULL)
```

Import expenses data

```{r}
expense_files <-
dir(path = expenses_path, 
    pattern = "^Transactions.*", 
    all.files = FALSE, 
    full.names = TRUE, 
    recursive = FALSE, 
    ignore.case = TRUE)

expenses_df <-
expense_files %>% 
  map_df(.f = read_csv) %>% 
  mutate(Account = str_to_lower(Account)) %>% 
  mutate(Branch = if_else(str_detect(string = Account, pattern = "kikuyu|pizzeria"), true = "Kikuyu", false = "Kabete")) %>% 
  mutate(Date = dmy(Date)) %>% 
  arrange(Branch,
          desc(Date))

expenses_df

expenses_df %>%
  count(Branch)

expenses_df %>% 
  group_by(Branch) %>% 
  summarise(range(Date))
```

Read salaries and rent

```{r}
fixed_costs <-
read_csv(file = paste0(expenses_path,"fixed_costs.csv")) %>% 
  mutate(Date = paste0("01",Month,Year)) %>% 
  mutate(Date = dmy(Date)) %>% 
  select(-Month, -Year)

fixed_costs %>% count(Branch)
range(fixed_costs$Date)
```


```{r}
# Bind fixed and variable costs
all_expenses <- bind_rows(expenses_df, fixed_costs) %>% 
  mutate(Amount = abs(Amount)) %>% 
  arrange(Branch,
          desc(Date))

all_expenses <-
all_expenses %>% 
  mutate(Category = str_replace(string = Category, pattern = "Salary", replacement = "Salaries"))
```

Simple summaries for expenses.

```{r}

expenses_summary <-
all_expenses %>% 
  group_by(Branch, Year = year(Date)) %>% 
  mutate(expenses_yr = sum(Amount, na.rm = TRUE)) %>% 
  ungroup() %>% 
  group_by(Branch, Year, Month = month(Date)) %>% 
  mutate(expenses_mon = sum(Amount, na.rm = TRUE)) %>% 
  ungroup() %>% 
  select(Branch, Year, Month, expenses_mon, expenses_yr) %>% 
  group_by(Branch, Year, Month) %>% 
  summarise(expenses_mon = mean(expenses_mon),
            expenses_yr = mean(expenses_yr)) %>% 
  arrange(Branch, Year, Month) %>% 
  ungroup()

expenses_summary
```

Sales, expenses, and profit summaries

```{r}
profit_summary <- 
  inner_join(x = sales_summary, 
             y = expenses_summary, by = c("Branch","Year","Month")) %>% 
  mutate(profit_mon = revenue_mon - expenses_mon)

profit_summary %>% 
  select(Branch, Year, Month, revenue_mon, expenses_mon, profit_mon) %>% 
  filter(Year == 2021, Branch == "Kikuyu")
```

Exporting current data

```{r}
# Export data
exp_file <- paste0("CF Data ",as.character(Sys.time()),".xlsx")
exp_file <- gsub(pattern = ":|-",replacement = "",x = exp_file)

write_xlsx(x = list(Sales = sales_data, 
                    Costs = all_expenses), 
           path = paste0(main_path,exp_file))
```

Merge with all previous dashboard data

```{r}
this_sales <- read_excel(paste0(main_path,exp_file), sheet = "Sales") %>% 
  mutate(Date = as.Date(Date))

this_costs <- 
  read_excel(paste0(main_path,exp_file), sheet = "Costs") %>% 
  mutate(Date = as.Date(Date)) %>% 
  select(Date,Branch,Category,Amount) %>% 
  rename(Item = Category,
         "Total cost" = Amount)

main_sales <- read_excel(paste0(main_path,"Crafted Pizza Data.xlsx"), sheet = 'Sales') %>% 
  mutate(Price = as.double(Price),
         Date = as.Date(Date))

main_costs <- read_excel(paste0(main_path,"Crafted Pizza Data.xlsx"), sheet = 'Costs') %>% 
  mutate(Date = as.Date(Date))

all_sales <- bind_rows(main_sales,this_sales) %>% 
  mutate(Branch = coalesce(Branch,"Kikuyu")) %>% 
  arrange(Date)

all_costs <- bind_rows(main_costs,this_costs) %>% 
  mutate(Branch = coalesce(Branch,"Kikuyu")) %>% 
  arrange(Date)

main_coords <- read_excel(paste0(main_path,"Crafted Pizza Data.xlsx"), sheet = 'Coordinates') 
```

Export merged data

```{r}
all_sales <- all_sales %>% arrange(desc(Date), Branch)

all_costs <- all_costs %>% arrange(desc(Date), Branch)
```

Monthly and daily summaries.

Summarise costs/sales by day then do an full_join.

```{r}
monthly_sales <-
all_sales %>% 
  select(Date, Revenue, Branch) %>% 
  group_by(Date, Branch) %>% 
  summarise(Revenue = sum(Revenue))

monthly_costs <-
all_costs %>% 
  select(Date, `Total cost`, Branch) %>% 
  group_by(Date, Branch) %>% 
  summarise(`Total cost` = sum(`Total cost`))

daily_summaries <-
monthly_sales %>% 
  full_join(monthly_costs, by = c("Date", "Branch")) %>% 
  replace_na(list(Revenue = 0, `Total cost` = 0)) %>% 
  mutate(profit = Revenue - `Total cost`) %>% 
  mutate(year_month = paste0(year(Date), strftime(x = Date, format = "%m"))) %>% 
  relocate(year_month, .after = Date) %>% 
  arrange(Branch, desc(Date)) %>% 
  ungroup()

monthly_summaries <-
daily_summaries %>% 
  select(-Date) %>% 
  group_by(year_month, Branch) %>% 
  summarise(across(.cols = c(Revenue, `Total cost`, profit), .fns = sum)) %>% 
  arrange(Branch, desc(year_month)) %>% 
  ungroup()
```

Creating key for joining the daily summaries and monthly summaries.

```{r}
monthly_summaries <-
monthly_summaries %>% 
  mutate(key = paste0(Branch,"_",year_month))

daily_summaries <-
daily_summaries %>% 
  mutate(key = paste0(Branch,"_",year_month))
```


Sanity checks...

```{r}
monthly_summaries %>% filter(profit < -1000)

monthly_summaries %>% 
  mutate(year = str_sub(string = year_month, start = 1, end = 4)) %>% 
  group_by(Branch, year) %>% 
  summarise(sum(profit))
```

Power BI file.

```{r}
now <- Sys.time()
now <- gsub(pattern = ":|-", replacement = "", x = now)
now_file <- paste0("CF_PowerBI_Data_",now,".xlsx")

write_xlsx(x = list(Daily_Summaries = daily_summaries,
                    Monthly_Summaries = monthly_summaries,
                    Sales = all_sales, 
                    Costs = all_costs, 
                    Coordinates = main_coords), 
           path = paste0(main_path,now_file))
```

====== END ===========

```{r}
daily_summaries %>% 
  group_by(Branch) %>% 
  filter(year(Date)==2021) %>% 
  summarise(across(.cols = c(Revenue, `Total cost`, profit), .fns = sum))
```



```{r}
sales_data %>% 
  mutate(month = month(Date),
         year = year(Date)) %>% 
  group_by(Category,year,month) %>% 
  summarise(revenue = sum(Revenue),
            n = n()) %>%
  summarise(revenue = mean(revenue),
            units_pm = mean(n)) %>% 
  filter(year != 2021)
```

